---
title: 'Lab 10 : Detecting Patters with Regular Expressions'
---

## Learning objectives

-   How to use regular expressions in R

## Load libaries

```{r}
library(tidyverse)
```

## What Are Regular Expressions?

Regular expressions are sequences of characters that define search patterns. In R, they are commonly used for:

- Searching text
- Extracting substrings
- Replacing patterns
- Validating formats (e.g., emails, dates)


### Key Functions in R for Regex

R provides several base functions that support regular expressions:

#### 1. `grep()` / `grepl()`
- **`grep()`** returns indices of matches.
- **`grepl()`** returns a logical vector indicating matches.

```{r}
grepl("cat", c("cat", "dog", "catalog"))
```

#### 2. `sub()` / `gsub()`
- **`sub()`** replaces the first match.
- **`gsub()`** replaces all matches.

```{r}
gsub("dog", "cat", "dog and dog")
```

#### 3. `regexpr()` / `gregexpr()`
- Return the position of the first/all matches and their lengths.

#### 4. `regmatches()`
- Extracts matched substrings based on `regexpr()` or `gregexpr()`.


### Regex Syntax Basics

Here are some commonly used regex symbols:

| Symbol | Meaning                          | Example         |
|--------|----------------------------------|-----------------|
| `.`    | Any character except newline     | `"a.b"` matches "acb", "a1b" |
| `^`    | Start of string                  | `"^cat"` matches "catfish" |
| `$`    | End of string                    | `"cat$"` matches "bobcat" |
| `*`    | 0 or more repetitions            | `"ca*t"` matches "ct", "cat", "caaaat" |
| `+`    | 1 or more repetitions            | `"ca+t"` matches "cat", "caaaat" |
| `?`    | 0 or 1 repetition                | `"ca?t"` matches "ct", "cat" |
| `[]`   | Character class                  | `"[cd]og"` matches "dog", "cog" |
| `|`    | OR                               | `"cat|dog"` matches either |
| `()`   | Grouping                         | `"(cat|dog)s?"` matches "cat", "cats", "dog", "dogs" |
| `\\`   | Escape special characters        | `"\\."` matches a literal dot |

###  Example Use Case

```{r}
text <- c("apple", "banana", "cherry", "date")
grep("a", text, value = TRUE)  # Matches strings containing 'a'
grep("^a", text, value = TRUE)  # Matches strings starting with 'a'
grep("a$", text, value = TRUE)  # Matches strings ending with 'a'
```

## Introduction to `stringr`

The base R regular expressions are great to use if you already use regular expressions in another programming language. Starting out the **`stringr`** package, part of the **tidyverse**, provides a cohesive and consistent set of functions for string manipulation. It simplifies working with regular expressions by offering:

- **Consistent function names** (`str_*`)
- **Predictable argument order** (string first, pattern second)
- **Built-in support for vectorized operations**
- **Better integration with tidyverse workflows**


### Why Use `stringr` rather than base R regex functions?

- **Cleaner syntax**: Easier to read and write than base R regex functions.
- **Tidyverse-friendly**: Works well with `dplyr`, `purrr`, and other packages.
- **Consistent behavior**: Avoids quirks of base R functions like `grep()` and `sub()`.


### Common `stringr` Functions for Regex

Here are some of the most useful functions when working with regular expressions:

#### 1. `str_detect()`
Checks if a pattern exists in a string.

```{r}
str_detect(c("apple", "banana", "cherry"), "^a")
```

#### 2. `str_replace()` / `str_replace_all()`
Replaces the first or all occurrences of a pattern.

```{r}
str_replace("cat and dog", "dog", "mouse")
```

```{r}
str_replace_all("cat and dog and dog", "dog", "mouse")

```

#### 3. `str_extract()` / `str_extract_all()`
Extracts the first or all matches of a pattern.

```{r}
str_extract("My email is test@example.com", "[a-z0-9._%+-]+@[a-z0-9.-]+\\.[a-z]{2,}")
```

#### 4. `str_match()` / `str_match_all()`
Extracts matched groups using parentheses.

```{r}
str_match("Name: John", "Name: (\\w+)")
```

#### 5. `str_split()`
Splits strings based on a pattern.

```{r}
str_split("apple,banana,cherry", ",")
```

### Regex Integration

All `stringr` functions accept regular expressions by default. You can use:

- `^`, `$`, `.`, `*`, `+`, `?`, `[]`, `()`, `|` — standard regex symbols
- Escaped characters like `\\d`, `\\s`, `\\w` for digits, whitespace, and word characters

## Exercises

[R for Data Science Chapter 3](https://r4ds.hadley.nz/data-transform).

Today we will walk through Chapter 15 Regular expressions in R for Data Science. As we did last week, by putting the examples and exercises in our own Quarto Markdown file, we can create own personal path through the Chapter.

### What to upload to Canvas

After you `Render` the qmd file to an html file, export the file to your computer and upload it to Canvas.
