---
title: "PB seminar"
---

First make a new folder "PBseminar" for the course materials



In RStudio Upload the file into your directory (import directly from github).

```{r}

raw_url <- "https://raw.githubusercontent.com/jeffreyblanchard/PathoGen2025/refs/heads/main/labs/NEON_soilMAGs_soilChem.csv"

mag <- readr::read_csv(raw_url, show_col_types = FALSE)

```

In co-pilot paste this prompt
This is a file with metagenomic assembled genomes (MAGs) from National Ecological Observatory Network metagenomic data.  It contains the MAG taxonomy, genome statistics and associated soil chemistry data. Using R and tidyverse could you write the code to make a graph shows the phyla that are present at the NEON sites.

```{r}

library(tidyverse)
library(janitor)    # clean_names()
library(lubridate)

# 1) Load
path <- "NEON_soilMAGs_soilChem.csv"
mag <- readr::read_csv(path, show_col_types = FALSE) |> janitor::clean_names()

# 2) (Optional) Define a quality filter for MAGs — adjust thresholds to taste
mag <- mag |>
  mutate(
    bin_quality = factor(bin_quality, levels = c("HQ","MQ","LQ"), ordered = TRUE),
    hq_flag = (bin_completeness >= 90 & bin_contamination <= 5) | bin_quality == "HQ"
  )

# 3) Choose whether to use all MAGs or only high-quality ones
use_hq_only <- TRUE
mag_use <- if (use_hq_only) mag |> filter(hq_flag) else mag

# 4) Compute presence per (site_id, phylum)
presence_df <- mag_use |>
  filter(!is.na(site_id), !is.na(phylum)) |>
  distinct(site_id, phylum, bin_id) |>               # distinct MAGs
  summarize(present = n() > 0, .by = c(site_id, phylum)) # logical presence

# 5) Plot presence/absence as a tile heatmap
ggplot(presence_df, aes(x = site_id, y = phylum, fill = present)) +
  geom_tile(color = "grey80") +
  scale_fill_manual(values = c(`TRUE` = "#2c7fb8", `FALSE` = "white"),
                    breaks = c(TRUE, FALSE),
                    labels = c("Present", "Absent")) +
  labs(title = "Phyla present at each NEON site",
       x = "Site", y = "Phylum", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

```

